---
title: "Applied Econometrics for Macroeconomics and Finance"
output:
  pdf_document: default
  html_document: default
date: 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```



```{r}
# Set the working directory
setwd("C:/Users/Baha/Downloads")
#load the required packages
library(astsa)
library(forecast)
library(timeSeries)
library(tidyverse)
#Load the dataset
report1 <-read.csv("report1.csv")
head(report1)
```


```{r}
#Check for missing values in the dataset
colSums(is.na(report1))
dim(report1)
#The dataset has missing values, we remove them.
report1 <-report1 %>% 
  na.omit()
dim(report1)
colSums(is.na(report1))
head(report1)
```


```{r}
#Inspect the structure of the cleaned dataset
str(report1)
```


# Model selection

For model selection, we'll use techniques like ACF/PACF plots, differencing, and information criteria (e.g., AIC, BIC) to choose appropriate ARIMA models for inflation.

```{r}
# ACF and PACF plots
acf(report1$INF, main="ACF Plot of Inflation")
pacf(report1$INF, main="PACF Plot of Inflation")
# Differencing if necessary
diff_inflation <- diff(report1$INF)
# ACF and PACF plots for differenced data
acf(diff_inflation, main="ACF Plot of Differenced Inflation")
pacf(diff_inflation, main="PACF Plot of Differenced Inflation")
```


```{r}
# Fit the first ARIMA model
model1 <- Arima(report1$INF,order = c(1,1,2))
# Display the fitted model
summary(model1)
```


```{r}
#Fit the second ARIMA model
model2 <-Arima(report1$INF, order = c(1,2,1))
#Print the summary of the model2
summary(model2)
```


```{r}
#Fit the third ARIMA model
model3 <-Arima(report1$INF, order = c(2,1,1))
summary(model3)
```

#Forecasting

Once we have selected three ARIMA models, we can use them to forecast inflation for 2023 and 2024.

```{r}
# Forecasting inflation for 2023 and 2024 using the selected models
#Forecasts for the first model
forecast_model1 <- forecast(model1, h=8)
# View the forecasts
print(forecast_model1)
```


```{r}
#Forecasts for the second model
forecast_model2 <- forecast(model2, h=8)
# View the forecasts
print(forecast_model2)
```


```{r}
#Forecasts for the third model
forecast_model3 <- forecast(model3, h=8)
# View the forecasts
print(forecast_model3)
```


#Comparison of the forecasted values for the three fitted models and the actual values of inflation

```{r}
# Actual inflation data for Q1 to Q3 in 2023
actual_inflation = c(5.6, 3.2, 4.8)
# Forecast values from each ARIMA model for Q1 to Q3 in 2023
forecast_inflation1 = c(4.9, 5.4, 4.9) 
forecast_inflation2 = c(4.0, 4.4, 4.2)
forecast_inflation3 = c(4.5, 4.7, 4.8)
# Calculate forecast errors for each model
forecast_errors <- function(actual, forecast) {
  errors <- actual - forecast
  mae <- mean(abs(errors))
  mse <- mean(errors^2)
  rmse <- sqrt(mse)
  return(c(mae, mse, rmse))
}

# Calculate forecast errors for each model
errors_model1 <- forecast_errors(actual_inflation, forecast_inflation1)
errors_model2 <- forecast_errors(actual_inflation, forecast_inflation2)
errors_model3 <- forecast_errors(actual_inflation, forecast_inflation3)
# Print forecast errors for each model
cat("Model 1:\n")
cat("MAE:", errors_model1[1], "\n")
cat("MSE:", errors_model1[2], "\n")
cat("RMSE:", errors_model1[3], "\n\n")

cat("Model 2:\n")
cat("MAE:", errors_model2[1], "\n")
cat("MSE:", errors_model2[2], "\n")
cat("RMSE:", errors_model2[3], "\n\n")

cat("Model 3:\n")
cat("MAE:", errors_model3[1], "\n")
cat("MSE:", errors_model3[2], "\n")
cat("RMSE:", errors_model3[3], "\n")
```
# Model 3 #####


#Term structure of interest rates

```{r}
#Evidence of whether the interest rates are stationary
library(urca)
# Perform Augmented Dickey-Fuller (ADF) test for each interest rate
adf_test_rt <- ur.df(report1$OCR, type = "drift", lags = 4)
adf_test_r3 <- ur.df(report1$r3Mth, type = "drift", lags = 4)
adf_test_r5 <- ur.df(report1$r5Yr, type = "drift", lags = 4)
adf_test_r10 <- ur.df(report1$r10Yr, type = "drift", lags = 4)

# Print ADF test results for each interest rate
print("ADF Test Results for rt:")
summary(adf_test_rt)
print("ADF Test Results for r3:")
summary(adf_test_r3)
print("ADF Test Results for r5:")
summary(adf_test_r5)
print("ADF Test Results for r10:")
summary(adf_test_r10)
```



```{r}
# Define the variables for cointegration analysis
interest_rates <- report1[, c("OCR", "r3Mth", "r5Yr", "r10Yr")]

# Perform Johansen cointegration test
johansen_test <- ca.jo(interest_rates, type = "trace", ecdet = "const", K = 4)

# Print the results
summary(johansen_test)
```
#Stationarity Test for Spreads

```{r}
# Calculate the interest rate spreads
spread_3_1 <- report1$r3Mth - report1$OCR
spread_5_3 <- report1$r5Yr - report1$r3Mth
spread_10_5 <- report1$r10Yr - report1$r5Yr

# Perform ADF test for each spread
adf_test_3_1 <- ur.df(spread_3_1, type = "drift", lags = 4)
adf_test_5_3 <- ur.df(spread_5_3, type = "drift", lags = 4)
adf_test_10_5 <- ur.df(spread_10_5, type = "drift", lags = 4)

# Print ADF test results for each spread
print("ADF Test Results for Spread st,3-1:")
summary(adf_test_3_1)
print("ADF Test Results for Spread st,5-3:")
summary(adf_test_5_3)
print("ADF Test Results for Spread st,10-5:")
summary(adf_test_10_5)
```

#Dynamic Effects on Inflation

```{r}
# Load necessary packages
library(vars)

# Create a data frame with the spreads and inflation
Spreads_inflation <- data.frame(spread_3_1, spread_5_3, spread_10_5, inflation = report1$INF)

# Estimate Vector Auto regressive model
var_model <- VAR(Spreads_inflation, p = 2)  
# Perform Granger causality test to analyze dynamic effects
# Perform Granger causality test
granger_test <- causality(var_model, cause = c("spread_3_1", "spread_5_3", "spread_10_5"))

# Print Granger causality test results
print("Granger Causality Test Results:")
print(granger_test)
```


```{r}
# Compute impulse response function (IRF)
irf <- irf(var_model, impulse = "spread_3_1", response = "inflation", boot = FALSE, n.ahead = 10)
irf
# Extract IRF at different time horizons
irf_at_impact <- irf$irf[[1]][1, ]
irf_one_year <- irf$irf[[1]][2, ]
irf_ten_years <- irf$irf[[1]][11, ]

# Print the effects on inflation at different time horizons
print("Effects on Inflation:")
print("At the time of impact:")
print(irf_at_impact)
print("One year after impact:")
print(irf_one_year)
print("Ten years after impact:")
print(irf_ten_years)
```


```{r}
# Compute impulse response function (IRF)
irf <- irf(var_model, impulse = "spread_10_5", response = "inflation", boot = FALSE, n.ahead = 10)

# Extract IRF at different time horizons
irf_at_impact <- irf$irf[[1]][1, ]
irf_one_year <- irf$irf[[1]][2, ]
irf_ten_years <- irf$irf[[1]][11, ]

# Print the effects on inflation at different time horizons
print("Effects on Inflation:")
print("At the time of impact:")
print(irf_at_impact)
print("One year after impact:")
print(irf_one_year)
print("Ten years after impact:")
print(irf_ten_years)
```



```{r}
report2 <-read.csv("C:/Users/Baha/Downloads/report2 .csv")
#Check for missing values in the data
dim(report2)
sum(is.na(report2)) # No missing values in the data
head(report2)
```


```{r}
# Load necessary libraries
library(readr)
library(tseries)
library(rugarch)
# Extract the variables
rBHP <- report2$BHP
rCBA <- report2$CBA

# Define the GARCH model specification
specs <- ugarchspec(mean.model = list(armaOrder = c(0, 0)), 
                   variance.model = list(model = "sGARCH"), 
                   distribution.model = "std")

# Estimate GARCH model for rBHP
garch_rBHP <- ugarchfit(specs, data = rBHP, solver.control = list(trace = FALSE))

# Summary of the estimated model for rBHP
print(garch_rBHP)

# Estimate GARCH model for rCBA
garch_rCBA <- ugarchfit(specs, data = rCBA, solver.control = list(trace = FALSE))

# Summary of the estimated model for rCBA
print(garch_rCBA)
```


```{r}
# Extract the volatility estimates from the GARCH models
volatility_rBHP <- sigma(garch_rBHP)
volatility_rCBA <- sigma(garch_rCBA)

# Read the dataset containing pVIX_t
pVIX <- report2$VIX
# Plot the volatility series and pVIX_t for visual comparison
plot(volatility_rBHP, type = "l", col = "blue", ylim = c(0, max(max(volatility_rBHP), max(volatility_rCBA), max(pVIX))), 
     xlab = "Date", ylab = "Volatility", main = "Comparison of Volatility Estimates")
lines(volatility_rCBA, col = "red")
lines(pVIX, col = "green")
legend("topright", legend = c("rBHP", "rCBA", "pVIX"), col = c("blue", "red", "green"), lty = 1, cex = 0.8)

# Calculate summary statistics for comparison
summary(volatility_rBHP)
summary(volatility_rCBA)
summary(pVIX)
```


```{r}
# Calculate probability of returns less than 0.01% for specific dates
# Filter data for the specified dates
returns_28_02_2024 <- subset(report2, Date == "28/02/2024")$BHP
returns_29_02_2024 <- subset(report2, Date == "2024/02/29")$BHP
returns_01_03_2024 <- subset(report2, Date == "2024/03/01")$BHP

# Calculate the probability of returns less than 0.01%
prob_less_than_0.01_28_02_2024 <- mean(report2$BHP < 0.0001, na.rm = TRUE)
prob_less_than_0.01_29_02_2024 <- mean(report2$BHP < 0.0001, na.rm = TRUE)
prob_less_than_0.01_01_03_2024 <- mean(report2$BHP < 0.0001, na.rm = TRUE)

# Print the probabilities
cat("Probability of returns less than 0.01% on 2024-02-28:", prob_less_than_0.01_28_02_2024, "\n")
cat("Probability of returns less than 0.01% on 2024-02-29:", prob_less_than_0.01_29_02_2024, "\n")
cat("Probability of returns less than 0.01% on 2024-03-01:", prob_less_than_0.01_01_03_2024, "\n")
```


```{r}
# Filter data for the specified dates
returns_28_02_2024 <- subset(report2, Date == "28/02/2024")$CBA
returns_29_02_2024 <- subset(report2, Date == "2024/02/29")$CBA
returns_01_03_2024 <- subset(report2, Date == "2024/03/01")$CBA

# Calculate the probability of returns less than 0.01%
prob_less_than_0.01_28_02_2024 <- mean(report2$CBA < 0.0001, na.rm = TRUE)
prob_less_than_0.01_29_02_2024 <- mean(report2$CBA < 0.0001, na.rm = TRUE)
prob_less_than_0.01_01_03_2024 <- mean(report2$CBA < 0.0001, na.rm = TRUE)

# Print the probabilities
cat("Probability of returns less than 0.01% on 2024-02-28:", prob_less_than_0.01_28_02_2024, "\n")
cat("Probability of returns less than 0.01% on 2024-02-29:", prob_less_than_0.01_29_02_2024, "\n")
cat("Probability of returns less than 0.01% on 2024-03-01:", prob_less_than_0.01_01_03_2024, "\n")
```

